sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-29 doc/paper.Rmd
minor model description and pcod example edits

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.

@@

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to an equidistant projection such as UTMs such that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
<!-- Since this is not particular to your package, maybe the e.g. for how to do this should be moved to the appendix? I do explain the transformation for the owl data, but it could be added to the pcod appendix too? -->
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, BC, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

@@

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y` (km).
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

@@

We could extend our model to a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1).

@@

We could generate a population index that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \@ref(fig:pcod-fig)E, Appendix 2).

>>>>

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass a sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.

@@

An sdmTMB model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
<!-- Since this is not particular to your package, maybe the e.g. for how to do this should be moved to the appendix? I do explain the transformation for the owl data, but it could be added to the pcod appendix too? -->
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

@@

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

@@

We could extend this spatial model to be a spatiotemporal model simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured in time.
Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1).

@@

We could generate a population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \@ref(fig:pcod-fig)E, Appendix 2).
